A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-20 paper.tex
Attempt to smooth the narrative

\section*{Event ARGs}\label{eARG}
Genome ARGs model the inheritance between specific genomes,
annotating the edges of the graph with the intervals
of genetic inheritance. It is primarily concerned with
describing relationships and the details of inheritance
between pairs of genomes.
In contrast, the classical ARG data structure is defined
by evolutionary \emph{events},
and we therefore refer to it as an ``event ARG'' (eARG).
The difference between gARGs and eARGs is subtle and the distinctions
at times rather porous. In essense, our main goal of introducing the
term ``Event ARG'' is to contrast the flexibility of the genome ARG
view with current practice. It is therefore useful to consider
the definition of an ARG from recent work.

% Todo this is rough. Ideally we want a bit of commentary running through
% the quotations.
\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''
\citet{shipilina2023origin} emphasise the reality  ``The ARG describes the complete
ancestry of a sample of genomes through a series of real coalescence and
recombination events''. In contrast,
\citet{brandt2021evaluation} tie the definition of an ARG explicitly
to a generative process:
``The full ancestral recombination graph (ARG) is a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination.''
In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [FIXME]
The common features of these definitions is that they are expressed
in terms of evolutionary events, and emphasise the exhaustiveness
of the record (full, complete). We will examine the consequences of both
of these properties in different contexts throughout the remainder of
this article.

Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
% % Papers that do not discuss how positions work:
% % - Mathieson and Scally: no discussion of how it works concretely (but quite
% %   close to gARG, in some ways)
% %
% The common features of these definitions is that they are expressed
% in terms of evolutionary events, emphasise the exhaustiveness
% of the record (full, complete), and they do not specify the mechanism
% by which partial genetic inheritance occurs. Just as the graph topology
% in a gARG defines the paths along which genetic inheritance occurs, some
% additional information is required to define the actual regions of genome
% inherited.
Classically, this is done by having two different types
of node:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
% In a common ancestor event the inbound lineages are merged into a
% single ancestral lineage with one parent, whereas in a recombination
% event a lineage is split into two independent
% ancestral lineages with an associated breakpoint.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use a common convention here
that the order of parents
is significant, but additional information is sometimes incorporated to
distinguish left and right parents
\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

>>>>

\section*{Event ARGs}\label{eARG}
Genome ARGs are concerned with
describing the details of genetic inheritance
between specific pairs of genomes.
In contrast, ARGs are classically seem as being defined
by the evolutionary \emph{events} in the history of a sample.
The difference between gARGs and this prevalent view of
what defines an ARGs is subtle and the distinctions sometimes porous.
It is therefore useful to first survey some representative
definitions of
an ARG from the literature, to allow us to capture the common features
of these definitions
(see Appendix XXX for a full review).

\citet{brandt2021evaluation} follow the original Griffiths definition of
an ARG by coupling it directly to a generative process, stating
that the ``full ancestral recombination graph (ARG) is a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination.''
\citet{shipilina2023origin} emphasise the true existance of the
corresponding events (rather than the generative process) by saying that
``[t]he ARG describes the complete
ancestry of a sample of genomes through a series of real coalescence and
recombination events.''
Following this trend to decouple the concept of an ARG from
generative processes~\citep{minichiello2006mapping},
\cite{mathieson2020ancestry}
define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''
In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]

% Papers that do not discuss how positions work:
% - Mathieson and Scally: no discussion of how it works concretely (but quite
%   close to gARG, in some ways)
% TODO list out some of these as examples of where people don't bother
% getting into this important detail?
The common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``Event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use a common convention here
that the order of parents
% TODO list papers where people use ordering? ARGweaver I think?
is significant,
but additional information is sometimes incorporated to
distinguish left and right parents
\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
